Effectful Semantics for Substructural SSA

### 2023-11-07 pldi24/paper.tex
Tap...
Tap

We can now define the \textit{trace monad} of a stream action as follows:
\begin{definition}[Trace Monad]
  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A \times M + I)\), with
  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
  and 
  \[\mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}\]
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with iteration operator
  \(f^\dagger(a) = \bigcup_{n \in \nats \cup \{\infty\}}f_n(a)\), where
  \TODO{this}
\end{definition}

@@

\begin{definition}[Submonad] 
  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and, for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where \(f^\dagger\) is computed as in \(\mb{T}\).
\end{definition}
Some important examples of submonads include:
\begin{itemize}
  \item \todo{unit submonad of every monad}
  \item \todo{union, intersection of (Elgot) submonads; lattice of submonads}
  \item \todo{commuting absorbtive submonad of trace monad: example of...}
  \item \todo{central submonad. Go cite \cite{carette-central-2022}}
  \item \todo{deterministic, idempotent trace monad: relevant submonad}
  \item \todo{noneffectful trace monad: affine submonad}
  \item \todo{cap and centralize, cap and purify}
\end{itemize}
\TODO{in particular, these induce a substructural effectful category...}

>>>>

We can now define the \textit{trace monad} of a stream action as follows:
\begin{definition}[Trace Monad]
  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A \times M + I)\), with
  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
  and 
  \[\mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}\]
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with iteration operator
  \(f^\dagger(a) = f^\infty(a) \cup \bigcup_{n \in \nats}\{\iota_0 (\iota_0\;b, m) \in f_n(a)\} \cup \{\iota_1 t \in f_n(a)\}\), where
  \[
    f_0 = f, \qquad f_{n + 1} = f \gg [\upg{\iota_0}{}, f_n]
  \]
  \[
    f^\infty(a) = \{\Theta (\lambda n. m_n) \mid a_0 = a \land \forall n, \exists a_{n + 1}. \iota_0 (\iota_1\;a_{n + 1}, m_n) \in f(a_n)\}
  \]
  \TODO{clean this up}
\end{definition}

@@

\begin{definition}[Submonad] 
  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and, for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where \(f^\dagger\) is computed as in \(\mb{T}\).
\end{definition}
Some important examples of submonads include:
\begin{itemize}
  \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\); similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
  \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise unions \(\bigvee_i\mb{S}_i = \lambda A. \bigcup_i \mb{S}_i\;A\) and intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are submonads, forming a lattice with bottom element \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\)
  \item \todo{commuting absorbtive submonad of trace monad: example of...}
  \item \todo{central submonad. Go cite \cite{carette-central-2022}}
  \item \todo{deterministic, idempotent trace monad: relevant submonad}
  \item \todo{noneffectful trace monad: affine submonad}
  \item \todo{cap and centralize, cap and purify}
\end{itemize}
\TODO{in particular, these induce a substructural effectful category...}
